 

### 2023-04-26 markdown/methods.tex
first draft flowchart for analysis pipeline

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
2cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\allsectionsfont{\centering}
\subsectionfont{\raggedright}
\subsubsectionfont{\raggedright}

\pagenumbering{gobble}

\begin{centering}

\vspace{3cm}

\includegraphics[width=0.2\linewidth]{FHCW_logo} 
\includegraphics[width=0.2\linewidth]{KI_logo}

\vspace{1cm}

\Large
\doublespacing
{\bf Design and implementation of analysis pipeline for single cell type proteomics data}

\vspace{1 cm}

\normalsize
\singlespacing
By

\vspace{0.5 cm}

\Large

{\bf Lukas Gamp}

\vspace{1.5 cm}

in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics

\vspace{1.5 cm}

\normalsize
mm yy
\end{centering}

\newpage

\pagenumbering{gobble}

\begin{centering}

{\bf Abstract}

\end{centering}

\spacing{1.5}

(the spacing is set to 1.5)

no more than 250 words for the abstract

\begin{itemize}
\tightlist
\item
  a description of the research question/knowledge gap -- what we know
  and what we don't know
\item
  how your research has attempted to fill this gap
\item
  a brief description of the methods
\item
  brief results
\item
  key conclusions that put the research into a larger context
\end{itemize}

\pagenumbering{roman}

\newpage

\centering
\raggedright
\newpage
\tableofcontents

\newpage

\section*{Acknowledgements}

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{proteomics}{%
\subsection{Proteomics}\label{proteomics}}

The proteome is referred to the sum of all proteins of a given sample at
a given time. In the past several quantitative and qualitative assays
were used to enlighten the protein composition of a sample.

An early approach of qualitative analysis of the cellular proteome
involved labeling with fluorescent antibodies and imaging. The major
disadvantage of this technique was the limitation to only stain a few
proteins per cell. For quantification procedures such as single-cell
Western blots, immunoassays or CyTOF have been used. Other disadvantages
are the ability to permeate cells, accessibility and binding of the
epitope and the creation of specific antibodies for a given protein
\citep{Budnik2018}.

One of those techniques involved RNA-sequencing. Since RNA involves also
non-coding RNA, the amount of RNA is often not proportional to the
content of proteins in a cell. So the proteinaceous content of a cell
was only predicted and quantitative analysis was not possible.

\hypertarget{mass-spectrometry}{%
\subsection{Mass Spectrometry}\label{mass-spectrometry}}

Mass spectrometry enables qualitative and quantitative analysis of the
entire repertoire of a biological sample. The availability of gene
sequences in databases and the ability to match proteins against those
sequences with computational methods makes it possible to identify
alterations of a sample on a protein level. These alterations can rely
on the sequence level or could be to post-translational modifications
(PTMs) such as phosphorylation, methylation or else
\citep{Aebersold2003}.

Mass to charge ratio (m/z)

\hypertarget{cell-isolation}{%
\paragraph{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\paragraph{Lysis}\label{lysis}}

\hypertarget{digestion}{%
\paragraph{Digestion}\label{digestion}}

\hypertarget{ionization}{%
\paragraph{Ionization}\label{ionization}}

\hypertarget{coupled-mass-spectrometry-msms}{%
\paragraph{Coupled mass-spectrometry
(MS/MS)}\label{coupled-mass-spectrometry-msms}}

\hypertarget{labeling-techniques}{%
\paragraph{Labeling techniques}\label{labeling-techniques}}

@@

\hypertarget{isobaric-labeling}{%
\subparagraph{Isobaric labeling}\label{isobaric-labeling}}

Tandem mass tag (TMT)

@@

\hypertarget{ms-spectrum}{%
\subsubsection{MS-Spectrum}\label{ms-spectrum}}

@@

\hypertarget{interpretation-of-the-data}{%
\subsubsection{Interpretation of the
data}\label{interpretation-of-the-data}}

\hypertarget{data-acquisition}{%
\paragraph{Data acquisition}\label{data-acquisition}}

@@

\hypertarget{three-dimensional-peak-detection}{%
\subparagraph{Three-dimensional peak
detection}\label{three-dimensional-peak-detection}}

@@

\hypertarget{deisotoping}{%
\subparagraph{Deisotoping}\label{deisotoping}}

@@

\hypertarget{label-detection}{%
\subparagraph{Label detection}\label{label-detection}}

@@

\hypertarget{improving-peptide-mass-accuracy}{%
\subparagraph{Improving peptide mass
accuracy}\label{improving-peptide-mass-accuracy}}

@@

\hypertarget{peptide-search}{%
\subparagraph{Peptide search}\label{peptide-search}}

@@

\hypertarget{protein-assembly}{%
\subparagraph{Protein assembly}\label{protein-assembly}}

After these calculation the identified peptides can be aggregated
according to it`s respective protein and quantified. The mentioned
metrics indicating the performance of the peptide search can be used in
downstream analysis. A so called razor peptide indicates the group with
the highest number of identified peptides. Quantification is enabled by
taking only unique peptides into contrast. Posterior error
probabilities, which refers to the chance that the found peptide is a
random event, are multiplied and only distinct sequences with the
highst-scoring are used.

\hypertarget{data-processing}{%
\paragraph{Data processing}\label{data-processing}}

Further analysis can be done with R and respective packages such as
bioconductor. Since there is no state of the art established, analysis
varies upon experimental design.

\hypertarget{reading-the-data}{%
\subparagraph{Reading the data}\label{reading-the-data}}

After processing MaxQuant creates a directory containing all results as
.txt file. The evidence.txt file include all peptide sequence matches
(PSM) with their respective proteins and statistical parameters.

@@

In order to streamline the analysis of multiple experiments, object
oriented programming can be applied. The approach in R is to create a so
called Q-feature object, which contains all variables and metadata in a
hierarchical structure. The structure enables sub setting for further
analysis.

\hypertarget{filtering}{%
\paragraph{Filtering}\label{filtering}}

Peptides having a low abundance will be set to zero by the analysis. The
value zero would imply that the sample does not contain the respective
peptide, therefore replacing 0 with not applicable (NA) is crucial for
understanding and interpretation of MS-data.

@@

The author of the SCoPE2 \citep{Petelski2021} pipeline recommends
calculating the sample to carrier ration (SCR). This can be done by
dviding the ion intensity of the sample by the reporter ion intensity of
the carrier channel. However in our experiment no carrier and reference
channel is used.

@@

@@

\pagenumbering{arabic}

\newpage

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\vspace{0.5cm}

Analysis of the data was done with MaxQuant \citep{Cox2008}

\newpage

\hypertarget{results}{%
\section{Results}\label{results}}

Some more guidlines from the School of Geosciences.

This section should summarise the findings of the research referring to
all figures, tables and statistical results (some of which may be placed
in appendices). - include the primary results, ordered logically - it is
often useful to follow the same order as presented in the methods. -
alternatively, you may find that ordering the results from the most
important to the least important works better for your project. - data
should only be presented in the main text once, either in tables or
figures; if presented in figures, data can be tabulated in appendices
and referred to at the appropriate point in the main text.

\textbf{Often, it is recommended that you write the results section
first, so that you can write the methods that are appropriate to
describe the results presented. Then you can write the discussion next,
then the introduction which includes the relevant literature for the
scientific story that you are telling and finally the conclusions and
abstract -- this approach is called writing backwards.}

\newpage

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

the purpose of the discussion is to summarise your major findings and
place them in the context of the current state of knowledge in the
literature. When you discuss your own work and that of others, back up
your statements with evidence and citations. - The first part of the
discussion should contain a summary of your major findings (usually 2 --
4 points) and a brief summary of the implications of your findings.
Ideally, it should make reference to whether you found support for your
hypotheses or answered your questions that were placed at the end of the
introduction. - The following paragraphs will then usually describe each
of these findings in greater detail, making reference to previous
studies. - Often the discussion will include one or a few paragraphs
describing the limitations of your study and the potential for future
research. - Subheadings within the discussion can be useful for
orienting the reader to the major themes that are addressed.

\newpage

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The conclusion section should specify the key findings of your study,
explain their wider significance in the context of the research field
and explain how you have filled the knowledge gap that you have
identified in the introduction. This is your chance to present to your
reader the major take-home messages of your dissertation research. It
should be similar in content to the last sentence of your summary
abstract. It should not be a repetition of the first paragraph of the
discussion. They can be distinguished in their connection to broader
issues. The first paragraph of the discussion will tend to focus on the
direct scientific implications of your work (i.e.~basic science,
fundamental knowledge) while the conclusion will tend to focus more on
the implications of the results for society, conservation, etc.

\newpage

\bibliography{bibliography}

\newpage

\hypertarget{appendixces}{%
\section{Appendix(ces)}\label{appendixces}}

\hypertarget{appendix-a-additional-tables}{%
\subsection{Appendix A: additional
tables}\label{appendix-a-additional-tables}}

Insert content for additional tables here.

\newpage

\hypertarget{appendix-b-additional-figures}{%
\subsection{Appendix B: additional
figures}\label{appendix-b-additional-figures}}

Insert content for additional figures here.

\newpage

\hypertarget{appendix-c-code}{%
\subsection{Appendix C: code}\label{appendix-c-code}}

Insert code (if any) used during your dissertation work here.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\hypertarget{cell-isolation}{%
\subsubsection{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\subsubsection{Lysis}\label{lysis}}

\hypertarget{digestion}{%
\subsubsection{Digestion}\label{digestion}}

\hypertarget{labeling-techniques}{%
\subsubsection{Labeling techniques}\label{labeling-techniques}}

@@

\hypertarget{isobaric-labeling}{%
\paragraph{Isobaric labeling}\label{isobaric-labeling}}

\hypertarget{tandem-mass-tag-tmt}{%
\subparagraph{Tandem mass tag (TMT)}\label{tandem-mass-tag-tmt}}

@@

\hypertarget{mass-spectrometry}{%
\paragraph{Mass Spectrometry}\label{mass-spectrometry}}

\hypertarget{ionization}{%
\subparagraph{Ionization}\label{ionization}}

\hypertarget{ms.1}{%
\subparagraph{MS.1}\label{ms.1}}

\hypertarget{coupled-mass-spectrometry-msms-ms.2}{%
\subparagraph{Coupled mass-spectrometry (MS/MS) \&
MS.2}\label{coupled-mass-spectrometry-msms-ms.2}}

Acquisition of the data was done with MaxQuant \citep{Cox2008} software
package.

\hypertarget{ms-spectrum}{%
\paragraph{MS-Spectrum}\label{ms-spectrum}}

@@

@@

\hypertarget{three-dimensional-peak-detection}{%
\paragraph{Three-dimensional peak
detection}\label{three-dimensional-peak-detection}}

@@

\hypertarget{deisotoping}{%
\paragraph{Deisotoping}\label{deisotoping}}

@@

\hypertarget{label-detection}{%
\paragraph{Label detection}\label{label-detection}}

@@

\hypertarget{improving-peptide-mass-accuracy}{%
\paragraph{Improving peptide mass
accuracy}\label{improving-peptide-mass-accuracy}}

@@

\hypertarget{peptide-search}{%
\paragraph{Peptide search}\label{peptide-search}}

@@

\hypertarget{protein-assembly}{%
\paragraph{Protein assembly}\label{protein-assembly}}

After these calculation the identified peptides can be aggregated
according to it`s respective protein and quantified. The mentioned
metrics indicating the performance of the peptide search can be used in
downstream analysis. A so called razor peptide indicates the group with
the highest number of identified peptides. Quantification is enabled by
taking only unique peptides into contrast. Posterior error
probabilities, which refers to the chance that the found peptide is a
random event, are multiplied and only distinct sequences with the
highest-scoring are used.

\hypertarget{data-processing}{%
\subsubsection{Data processing}\label{data-processing}}

Further analysis is done with R and respective packages such as
bioconductor. Since there is no state of the art established, analysis
varies upon experimental design. The workflow of the analysis will be
processed in a so called pipeline streamlining the data through steps
where individual results can be observed in visualizations and indivdual
calculations will be adapted according to user demands and experimental
properties.

\includegraphics{methods_files/figure-latex/data_processing_pipeline_flowchart-1.pdf}

\hypertarget{reading-the-data}{%
\paragraph{Reading the data}\label{reading-the-data}}

After processing MaxQuant creates a directory containing all results as
.txt file. The evidence.txt file include all peptide to spectrum matches
(PSM) with their respective proteins and statistical parameters.

@@

In order to streamline the analysis of multiple experiments, object
oriented programming can be applied. The approach in R is to create a so
called Q-feature object, which contains all variables and metadata in a
hierarchical structure. The structure enables sub setting for further
analysis \citep{Vanderaa2021}.

\hypertarget{zero-values}{%
\paragraph{Zero values}\label{zero-values}}

Peptides with low abundance are often set to zero during analysis.
However, assigning a value of zero may incorrectly suggest that the
sample does not contain the respective peptide. Given that it is highly
unlikely for a biological cell of a comparable type and function to not
contain a particular protein, replacing the zero value with ``not
applicable'' (NA) is crucial for understanding and interpreting MS data.

\hypertarget{exclude-reverse-matchescontaminants}{%
\paragraph{Exclude reverse
matches/contaminants}\label{exclude-reverse-matchescontaminants}}

@@

\hypertarget{filter-according-to-precursor-ion-fraction-pif}{%
\paragraph{Filter according to precursor ion fraction
(PIF)}\label{filter-according-to-precursor-ion-fraction-pif}}

During mass spectrometry, the ions detected in MS1 are further
fragmented through collision during multiple MS runs. The resulting
product ions are derived from precursor ions (also known as mother ions
or parental ions). Contaminant peptides can co-migrate in this process
and can be distinguished by the lower fraction of their respective
precursor ions \citep{Tannous2020}. These peptides need to be filtered
out during the analysis pipeline. A cutoff value, referenced in the
SCoPE2 pipeline \citep{Specht2021}, is applied in the user interface,
but it can be adjusted according to the needs of the biologist.

\hypertarget{filter-by-q-value}{%
\paragraph{Filter by q-value}\label{filter-by-q-value}}

@@

\hypertarget{peptide-spectrum-match-psm-aggregation-to-peptides}{%
\paragraph{Peptide spectrum match (PSM) aggregation to
peptides}\label{peptide-spectrum-match-psm-aggregation-to-peptides}}

In data science, aggregation refers to a row-wise operation that merges
data based on a particular column using a specific function. In the
context of processing from peptides to spectrum matches, the desired
column is the peptide sequence. To account for different distributions
across multiple assays, the median of the channel is used as the
function to aggregate multiple matches into one.

\hypertarget{join-assays-when-observing-multiple-comparable-batches-at-once}{%
\paragraph{Join assays when observing multiple comparable batches at
once}\label{join-assays-when-observing-multiple-comparable-batches-at-once}}

Sample size is often a limiting factor in hypothesis testing. A strict
quality control and the fact that TMT reagents are only available up to
18-plex can reduce the number of observed samples below the critical
threshold, leading to an early end of analyses. To overcome this
limitation, the provided software is capable of processing multiple runs
simultaneously, allowing for testing of multiple batches and increasing
the number of samples that can be included in the analysis.

\hypertarget{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}{%
\paragraph{Calculate reporter ion intensity (RI) and filter according to
median
RI}\label{calculate-reporter-ion-intensity-ri-and-filter-according-to-median-ri}}

Columns which do not meet the desired intensity can be filtered by a
threshold set on the RI. The median RI can also be used to check if an
entire channel has a lower detection level. This can be due two reasons.
One is the expression level of the given proteins in a cell. Meaning,
that the expression of the observed cell type is simply lower than the
other type. Another one could be a spillage of TMT detection in other
channels due incorrect or missing correction of the TMT isotopes.

\hypertarget{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}{%
\paragraph{Calculate and filter according to median coefficient of
variation (CV) per
cell/channel}\label{calculate-and-filter-according-to-median-coefficient-of-variation-cv-per-cellchannel}}

Depending having a bulk sample or single-cell sample choosing a minimum
of observed peptides and a cutoff value for the CV, changes the level of
confidence in the peptide data. The coefficient of variation of a
peptide is considered as the ratio of the standard deviation to the mean
and describes the relationship of the observed peptide signal over
multiple proteins (=razor proteins). Peptides having a high coefficient
of variation over many razor proteins are considered as noise and need
to be filtered out before statistical analysis.

\hypertarget{remove-peptides-with-high-missing-rate}{%
\paragraph{Remove peptides with high missing
rate}\label{remove-peptides-with-high-missing-rate}}

Although missing value imputation can be performed during the analysis
of multiple batches, peptides with missing detections across channels
can be problematic for quantification. The proteomic composition of a
biological sample is similar between replicates and even across groups.
However, the threshold of missingness (described as a fraction of the
row) can be set in the user interface and adjusted to enable different
experimental designs.

\hypertarget{aggregation-of-peptides-to-proteins}{%
\paragraph{Aggregation of peptides to
proteins}\label{aggregation-of-peptides-to-proteins}}

Similar to the already explained previous aggregation step the peptides
will be further processed into their respective proteins after the
quality control on the peptide level is performed.

\hypertarget{transformation-of-protein-expression-data}{%
\paragraph{Transformation of protein expression
data}\label{transformation-of-protein-expression-data}}

Data transformation applies a function to each value of a matrix or
array, so that: \[
y_i = f(x_i)
\] Depending on the distribution of the values in the observed
expression set, different transformations can be applied to fulfill
dependencies for statistical testing. In the shiny application various
procedures are implemented and can be further expanded upon request from
the user. The macrophage analysis done by Specht et. Al mentioned in
their SCoPE2 publication citep\{Specht2021\} uses the logarithm to the
base 2 to spread a compacted distribution and the remove skewness in the
dataset. This transformation was used as a reference when comparing
methods. However using the logarithm to the base 10 can be easy
interpretable way of defining expression data and will also be
facilitated by other transformation methods such as the boxcox method
which is also implemented in the application. The boxcox transformation
developed 1964 by Box and Cox citep\{Sakia1992\} applies a recursion on
the data by determining the parameter lambda. Depending on the size of
lambda a certain function will be applied on each value of the
expression set in order to introduce normality. As already mentioned
transformations applied by the BoxCox method range from log10 to taking
the square root of each value

@@
